openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-09-30 paper/policar2020-opentsne.tex
Edit methods section

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

@@

\usepackage{amsmath,amssymb}
\usepackage{booktabs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\newcommand{\opentsne}{\textsf{openTSNE}}
\endlocaldefs

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract}
Point-based visualisations of large, multi-dimensional data from molecular
	biology can reveal meaningful clusters. One of the most popular
	techniques to construct such visualisations is t-distributed stochastic
	neighbor embedding (t-SNE), for which a number of extensions have
	recently been proposed to address issues of scalability and the quality
	of the resulting visualisations. We introduce \opentsne, a modular
	Python library that implements the core t-SNE algorithm and its
	extensions. The library is orders of magnitude faster than existing
	popular implementations, including those from scikit-learn. Unique to
	\opentsne\ is also the mapping of new data to existing embeddings,
	which can surprisingly assist in solving batch effects.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{t-SNE}
\kwd{embedding}
\kwd{visualization}
\kwd{dimensionality reduction}
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, in particular for methods that produce informative data visualizations. Popular approaches include 
principal component analysis (PCA),
multidimensional scaling, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. 
Among these, t-SNE lately received much attention as it can address high
volumes of data and reveal meaningful clustering structure. Increased
resolution and throughput of modern molecular assays has lead to frequent use
of t-SNE in diverse fields including, but not limited to, single-cell
transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}.

Reports on single-cell gene expression data, our running example,
often start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}The figure shows the progression of t-SNE
	embeddings following theoretical advances, which are all included in
	\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
	single-cell gene-expression profiles of mouse retinal cells from the
	study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly,
	the data in \textbf{(c)} contains 27,499 expression profiles from mouse
	retinal cells from Shekhar \textit{et
	al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
	embedding following the parameter choices from the original publication
	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
	shows no preservation of the global organization of clusters,
	resulting from random initialization and an affinity model focused on
	preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE
	embedding, utilizing the latest theoretical advances and practical
	recommendations. The embedding is constructed using a multi-scale
	affinity model, preserving both short-range and long-range interactions
	between data points, and is initialized such that the global layout is
	meaningful. Unlike in \textbf{(a)}, the green and blue clusters
	representing different sub-types of amacrine and bipolar cells are now
	localized to the same regions of the space, indicating a higher level
	of similarity than to other cell types. \textbf{(c)} shows how existing
	t-SNE reference atlases can be used to embed new samples into existing
	embeddings. The positions of new data points correspond to cell types
	from the reference atlas.}
\end{figure*}

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new samples into existing
	embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
	well-defined clusters and uncover global relationships between clusters,
	and
\item improved defaults for learning rate and number of
	iterations~\cite{belkina2019automated} which produce embeddings at a low
	computational cost.
\end{enumerate}

The library is fully compatible with the Python data-science ecosystem (e.g.,
\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
and intuitive API to new users. Its modular design encourages
extensibility and experimentation with various settings and changes in the
analysis pipeline.

\opentsne\ is open-source and has become adopted by the wider community,
receeiving over 600 Github stars and averaging thousands of weekly downloads.
The source code is publicly available on Github at
\url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
\textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python
package managers.

Accessibility is one of the core design principles of \opentsne. Our aim is to
make the latest theoretical advances in t-SNE as accessible as possible to as
many users as possible. As such, our API is based on the familiar
\textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a
smooth transition from existing Python libraries. For example,
Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
\opentsne, each based on recent theoretical advances in the t-SNE algorithm. The code used
to generate these embeddings is shown below.

@@

\noindent The code snippet takes as input a \textsf{numpy} array or
\textsf{scipy} sparse matrix \texttt{X} and produces four
\texttt{TSNEEmbedding} objects. These may subsequently be subject to further
optimization under different parameter settings or may be used to embed new
samples into the embedding landscape. Additional parameters are omitted for
clarity but can be found in accompanying notebooks publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence that
every data set can accurately be described by using only two dimensions, and any such embedding will therefore inevitably lead to loss of
information. It is thus beneficial to examine multiple embeddings, each
of which provides a different perspective into topology and other
data characteristics.

@@

We illustrate this point by generating four embeddings of the same data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.  First,
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell subtypes.

When working with data containing millions of data points, standard t-SNE
embeddings often become unwieldy -- cluster boundaries become blurred, large
clusters absorb smaller ones, and relationships between clusters become
increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
profiles of over two million single cells captured at different time points in
mouse development. The embedding indicates numerous clusters with some
transition between time points as indicated by the color-coding, but this is
largely difficult to interpret. Kobak \& Berens observed that increasing
attractive forces between similar data points controlled via the
\textit{exaggeration} parameter leads to more compact clusters, and
subsequently more informative visualizations~\cite{kobak2019art}. For instance,
\ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the
overall structure of the data. Doubling the exaggeration again in
\ref{fig:cao}.c, we can now easily observe that the data is comprised of two
main groups of cells and eight somewhat smaller clusters. The visualization
also reveals a number of tiny clusters, possibly corresponding to
rare cell types.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}Increasing the exaggeration parameter produces more
  	compact clusters, which highlight the global organization of the clusters
  	and emphasizes continuities between cell states. The data set from Cao \textit{et
	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
	single cells. The data were collected from mice embryos at different
	developmental time-points at daily intervals after 9.5 to 13.5 days.
	\textbf{(a)} shows the data is comprised of two main components -- the
	neural tube and mesenchymal cells -- as well as several other smaller
	clusters. Additionally, the colors indicate developmental progression
	with red indicating least-developed cells and blue indicating most
	developed cells. The overall developmental trajectory is most apparent
	with higher levels of exaggeration, showing red cells slowly
	transitioning into blue cells. Progressively easing the exaggeration
	factor uncovers finer clusters within the larger groups as shown in
	\textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)},
	where we show the standard t-SNE with no exaggeration. 32,011 putative
	doublets are excluded from the visualizations.}
\end{figure*}

Exaggeration can highlight transitions between cell-states in
developmental studies. Standard t-SNE often produces embeddings with
clearly defined, discrete clusters. We have shown that we can adjust the level
of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
discrete clusters are often undesired in developmental studies where cells are assumed to follow a continuous trajectory between cell states. To this end, other
embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity between cell
states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time
points is difficult to observe due to a large number of sprawled out clusters.
On the other hand, the development becomes easier to follow when the
exaggeration factor is increased from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

Unlike other popular dimensionality reduction methods such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore embeddings of
new data points need to be found through the use of optimization
techniques~\cite{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data taken from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from
mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, 
we can correctly map to the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for
cell type classification in cases where cell identities are not known in
advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
embedding using labeled data from Hochgerner~\textit{et
al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
cells from the mouse brain. The authors assign a cell type to each cell, and
we can verify their accuracy by visualizing well-established gene markers for the
major cell types. We then embed cells from Harris~\textit{et
al.}~\cite{harris2018classes}, where labels are provided only for neuronal
cells. We quickly identify other non-neuronal cell types, including
oligodendrocytes and astrocytes. We use marker genes to validate that new cells
were correctly mapped into the reference landscape.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples
	into an existing, reference embedding. We construct a t-SNE embedding
	for the data from Hochgerner \textit{et
	al.}~\cite{hochgerner2018conserved} containing 24,185 developing,
	single cells from the mouse hippocampus. The data contains different
	types of neurons, supporting glia, and other vascular cells (upper
	left). Cell labels are taken from the original publications. The
	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
	hippocampal cells collected by Harris \textit{et
	al.}~\cite{harris2018classes} using the previous embedding as a
	reference. In their study, Harris \textit{et al.} collected 6,971
	single-cells and focused on identifying different types of inhibitory
	neurons. However, almost half of the collected cells are not neurons
	and were left uncharacterized. Inspecting where these cells are
	embedded in the reference embedding (bottom left) reveals that in
	addition to inhibitory neurons, the data contains several supporting
	glial cells as well as a small population of endothelial cells. We can
	verify the accuracy of our approach by inspecting marker genes for the
	major cell types in the reference (top row) and embedded samples
	(bottom row).}
\end{figure*}

These two examples demonstrate how \opentsne\ can be used to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing existing cell
atlases. This approach is general and not limited to single-cell gene
expression, but can be used with any tabular data set regardless of field.

@@

Versatility is another of  \opentsne s core design principles. Kobak \& Berens
recently introduced several recommendations and tricks to obtain better and
more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
of multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with larger data sets. \opentsne\ provides a flexible
API, where these improvements can be incorporated in just a few lines of code.
Furthermore, \opentsne\ supports custom affinity models, enabling users to
construct t-SNE embeddings on non-tabular data, e.g. relational data, as the
only affinity-model requirement is a notion of similarity between data points,
i.e. the distance matrix.  Finally, our comprehensive callback system can be
utilized to monitor and adapt different stages of the optimization phase and
has been used to construct visually appealing animations of the t-SNE
optimization process.

@@

One of t-SNE's common criticisms is limited scalability to large data sets
containing up to millions of data points~\cite{becht2019dimensionality}. These
criticisms are largely based on two factors. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was
exacerbated further by the fact that the most widely-used implementation of
t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes
performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is
perfectly capable of creating t-SNE embeddings of millions of data points in a
matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large
data sets in a matter of minutes.

>>>>

@@

\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\startlocaldefs
\newcommand{\opentsne}{\textsf{openTSNE}}
\endlocaldefs

@@

@@

@@

@@

\begin{abstractbox}
\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology can reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its
extensions. The library is orders of magnitude faster than existing
popular implementations, including those from scikit-learn. Unique to
\opentsne\ is also mapping new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

\begin{keyword}
\kwd{t-SNE}
\kwd{embedding}
\kwd{visualization}
\kwd{dimensionality reduction}
\kwd{Python library}
\end{keyword}

@@

\section*{Background}

The abundance of high-dimensional data sets in molecular biology calls for dimensionality reduction techniques, particularly for methods that produce informative data visualizations. Popular approaches include 
principal component analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much attention as it can address high volumes of data and reveal a meaningful clustering structure. Increased
resolution and throughput of modern molecular assays has lead to the frequent use of t-SNE in diverse fields including, but not limited to, single-cell transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\cite{sheth2019spatial} and metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape, where t-SNE embeds high-dimensional expression profiles into a two-dimensional
space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
\caption{\label{fig:macosko}
Three t-SNE embeddings demonstrate theoretical advances and corresponding implementations in
\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
single-cell gene-expression profiles of mouse retinal cells from Macosko \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)} contains 27,499 expression profiles from mouse
retinal cells from Shekhar \textit{et
al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
embedding following the parameter choices from the original publication
by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
shows no preservation of the global organization of clusters,
resulting from random initialization and an affinity model focused on
preserving local neighborhoods. \textbf{(b)} A modern t-SNE
embedding, utilizing the latest theoretical advances and practical
recommendations constructed using a multi-scale
affinity model, preserving both short-range and long-range interactions
between data points and initialized so that the global layout is
meaningful. Unlike in \textbf{(a)}, the green and blue clusters
representing different sub-types of amacrine and bipolar cells are now
localized to the same regions of the space, indicating a higher level
of similarity than to other cell types. Embedding in \textbf{(c)} shows how existing
t-SNE reference atlases can be used to place new samples into existing
embeddings. The positions of new data points correspond to cell types
from the reference atlas.
}
\end{figure*}

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions, including:
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new data samples into a fixed existing
	embeddings~\cite{policar2019embedding},
\item improved initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
	well-defined clusters and uncover global relationships between clusters,
	and
\item improved defaults for learning rate and number of
	iterations~\cite{belkina2019automated} which produce embeddings at a lower
	computational cost.
\end{enumerate}

The \opentsne\ is compatible with the Python data science ecosystem and libraries that include
\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}), thus providing a familiar and intuitive application program interface to the new users. Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
We have released \opentsne\ in an open-source to support adaptations by the wider community. At the time of the writing, the library averages several thousands of weekly downloads and has received almost 700 GitHub stars, an appreciation score gained by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

Accessibility of the latest theoretical advances in t-SNE is one of the core design principles of \opentsne. These advances are accessible through a programming interface of the library, where we closely follow the style of the interface as implemented in
\textsf{scikit-learn}~\cite{sklearn_api}, a popular library for machine learning. The following snippet exemplifies this interface and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the effect of a particular recent theoretical advance in the t-SNE algorithm.

@@

\noindent The code snippet considers a \textsf{numpy} array or
\textsf{scipy} sparse matrix \texttt{X} and creates four
\texttt{TSNEEmbedding} objects. These objects encode t-SNE embeddings that may subsequently be optimized under different parameter settings or used to embed new samples into the embedding landscape. Function calls may include additional parameters not shown in the example code but explained in accompanying notebooks available at \url{https://github.com/pavlin-policar/opentsne-paper}.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions. However, there is no evidence that every data set can accurately be described by using only two dimensions, and any such embedding will inevitably lead to a loss of information. Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.

@@

We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\cite{tasic2018shared}.  Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious. Embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters by the larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent the cluster absorption. We constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction between these two depictions reveals that many clusters are composed of subgroups representing different cell subtypes.

When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development. The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\cite{kobak2019art}. For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters. The visualization also reveals several tiny clusters, possibly corresponding to
rare cell types.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuity between cell states. The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells. The data were collected from mice embryos at different development stages at daily intervals after 9.5 to 13.5 days. \textbf{(a)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters. Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells. The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells. Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in
\textbf{(b)} with exaggeration of two and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration. 32,011 putative doublets are excluded from the visualizations.
}
\end{figure*}

Exaggeration can highlight transitions between cell states in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously. To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

Unlike other popular dimensionality reduction methods such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space. Therefore embeddings of new data points need to be found through optimization~\cite{policar2019embedding}. \opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape. The reference data from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases of unknown cell identities. For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain. The authors assign a type to each cell. We can verify their classification accuracy by visualizing the expression of well-established gene markers for the major cell types. We then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a constructed atlas. In Harris~\textit{et al.}~, labels are provided only for neuronal cells. In the resulting mapping, we can quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes. We can further use marker genes to validate that the mapping in the reference landscape was correct.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}
\openTSNE supports embedding new samples into an existing reference t-SNE landscape. For the series of visualizations shown in this figure, we first construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus. The data contains gene expression in different neurons, supporting glia, and other vascular cells (upper left). Data points representing cells are colored according to cell-types assigned in the original publication; see the legend from Fig.~\ref{fig:tasic} to map colors to cell-type. We then embed new, hippocampal cells collected in a study by Harris \textit{et al.}~\cite{harris2018classes} using the embedding of Hochgerner \textit{et al.} data as a reference. In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons. However, almost half of the collected cells are not neurons and were left uncharacterized. Inspecting the embeddings of these cells in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells. We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).
}
\end{figure*}

The examples presented above demonstrate how to use \opentsne\ to quickly gain insight into a newly-sequenced, single-cell data sets by utilizing existing cell atlases. The approach is general and not limited to single-cell gene expression, and one can, in principle, apply it to any tabular data set regardless of field.

@@

Versatility, the ability to use and combine different optimization approaches to construct the embedding spaces, is another of \opentsne's core design principles. Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}. These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with more massive data sets. \opentsne\ provides a flexible program interface to incorporate these improvements in just a few code lines. Furthermore, \opentsne\ supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity-model is the availability of similarity between data points, that is, the distance matrix.  Finally, the \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.

@@

One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\cite{becht2019dimensionality}. The culprit for slow response time stems from a particular optimization procedure and its specific implementation in the popular Python library. Namely, until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em e.g.} cells). The most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes when compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in \opentsne\, enabling the embedding of large data sets in a matter of minutes.
